---
title: "Dragen QC report for fastqc_metrics"
author: "Synnøve Yndestad"
date:  "`r format(Sys.Date())`"
output:
  html_document:
    theme: darkly
    highlight: zenburn
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Description

```{r, include=FALSE}
#  
#  Instructions for use:
#
#  
#  This workflow will search all folders and sub-folders in the working directory for *.fastqc_metrics.csv files created by Dragen.
#  It will then aggregate all fastqc metrics information and generate a QC report for all samples. 
#  Ideally it should be all samples from a single run.
#  
#   To use:
#   Place a folder containing the *fastqc_metrics.csv files from dragen output in the working directory.
#   The folder name will be assigned as Run-ID
#   Change the description of experiment below and knit report. 
#
```


 

**Change here to match sample description.**
I.e:

Breast cancer tumor biopsies from study "XX" 2021.  
RNAseq protocol: XX  
NovaSeq6000 output, processed by dragen v.3.8. 






```{r, echo =FALSE, message=FALSE, results='hide'}
library(tidyverse)
#Get directory , change Foldername as appropiate to str-remove folder name in filenames
dir <- gsub("\\./", "", list.dirs(recursive=FALSE)[-1]) %>% paste("/", sep= "")
#directory <- "dragen_fastqc/"
RunID <- gsub("\\./", "", list.dirs(recursive=FALSE)[-1])
```

# Report for dragen fastqc; RunID `r RunID`

Directory searched for *fastqc_metrics.csv files:
```{r, echo =FALSE, message=FALSE}
RunID
```


```{r, echo =FALSE, message=FALSE, results='hide'}
# read folder
library(tidyverse)
library(ggplot2)
library(plotly)


#Find all gene quantification files in the current folder and all subfolders and list file-names as tempFiles
tempFiles <- list.files(pattern="*fastqc_metrics.csv",recursive = TRUE)

#Import files as lists of dataframes
myList = sapply(tempFiles, read.csv, header = FALSE, simplify = FALSE, USE.NAMES = TRUE)
 
#inspect the first 10 rows in the first data frame in the list to see what kind of data you have
head(pluck(myList, 1), n = 10)  
```


```{r,  echo =FALSE, message=FALSE, results='hide'}
# Merge lists of dataframes, create an id column with sample name and select relevant columns and remove the list of data frames to save space.
All_files <- map_df(myList, ~as.data.frame(.x), .id="Sample")

# Optional, clean up sample names. Change if necessary.
All_files$Sample <- str_remove(All_files$Sample, dir)
All_files$Sample <- str_remove(All_files$Sample, ".fastqc_metrics.csv")
  
head(All_files, n= 4)

# Add Read1 Read" to sample name.
All_files$Sample <- paste(All_files$Sample, All_files$V2, sep = "_")
head(All_files, n= 4)
rm(myList)
```


```{r, echo = FALSE}
write_csv(All_files, paste0("Dragen_FastQC_aggregate_", RunID, ".csv"))
```

Aggregated output saved as "Dragen_FastQC_aggregate_RunID.csv"






## 1- Read Mean quality; Per-Sequence Quality Scores 
  
Total number of reads. Each average Phred-scale quality value is rounded to the nearest integer.

```{r, echo =FALSE, message=FALSE, results='hide'}
read_mean_quality  <- All_files %>% filter(V1 == "READ MEAN QUALITY")

read_mean_quality$V3 <- read_mean_quality$V3 %>% str_remove("Q") %>% 
                                                 str_remove(" Reads") %>% 
                                                 as.numeric()
```


```{r, echo =FALSE, message=FALSE}
p <-ggplot(data =read_mean_quality,  aes(x= V3, y=V4, group = Sample, color = Sample )) +
    labs(title ="Dragen-QC: Per-Sequence Quality Scores", y= "Count", x = "Mean Phred Quality Score") +
    geom_line() +
    theme(legend.position = "none")  +
   # theme(axis.text.x = element_text(angle=45)) +
    scale_y_continuous(labels = scales::label_comma() ) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
    theme_minimal()

ggplotly(p)%>% layout(showlegend = FALSE)
```



## 2- Positional Base Mean Quality; Per-Base Quality Scores

Average Phred-scale quality value of bases with a specific nucleotide at a given location in the read. Locations are listed first and can be either specific positions or ranges. The nucleotide is listed second and can be A, C, G, or T. N or ambiguous bases are assumed to have the system default value, usually QV2.

```{r, echo =FALSE, message=FALSE, results='hide'}
position_base_mean_quality  <- All_files %>% filter(V1 == "POSITIONAL BASE MEAN QUALITY")
head(position_base_mean_quality)
```



```{r, echo =FALSE, message=FALSE, results='hide'}
position_base_mean_quality <- position_base_mean_quality %>% separate(V3, c("Pos", "ReadPos", "Base", "AvgQ", "AvgQ2"))
position_base_mean_quality$ReadPos <- as.numeric(position_base_mean_quality$ReadPos)
head(position_base_mean_quality)
```

 
```{r, echo =FALSE, message=FALSE}
p <-ggplot(data =position_base_mean_quality,  aes(x= ReadPos, y=V4, color = Sample )) +
    labs(title ="Dragen-QC: Per-Base Quality Scores", y= "Avg Phred Quality Score", x = "Base position") +
    geom_line() +
    theme(legend.position = "none") +
    theme(axis.text.x = element_text(size=5, angle=45))+
    facet_wrap("Base") +
    theme_minimal()
   # theme(axis.text.x = element_text(angle=45)) +
   # scale_y_continuous(labels = scales::label_comma() ) +
    #scale_x_continuous(breaks = scales::pretty_breaks(n = 15))

ggplotly(p) %>% layout(showlegend = FALSE)
```



## 3- Positional Base Content

Number of bases of each specific nucleotide at given locations in the read. Locations are given first and can be either specific positions or ranges. The nucleotide is listed second and can be A, C, G, T, N.


Per-Position Sequence Content heatmap and Per-Position N Content

```{r, echo =FALSE, message=FALSE, results='hide'}
position_base_content <- All_files %>% filter(V1 == "POSITIONAL BASE CONTENT")
head(position_base_content)
position_base_content$V3 <- position_base_content$V3 %>% str_remove("ReadPos ") %>% 
                                                         str_remove(" Bases")
position_base_content <- position_base_content %>% separate(V3, c("ReadPos", "Bases"))
position_base_content$ReadPos <- as.numeric(position_base_content$ReadPos)
head(position_base_content)
```

 

```{r, echo =FALSE, message=FALSE, results='hide'}

#Now calculate % bases


position_base_content <-position_base_content %>% group_by(Sample, ReadPos) %>% 
                                                  mutate(Total = sum(V4)) 


position_base_content_w <- position_base_content  %>% pivot_wider(names_from = Bases,
                                       values_from = c(V4)) %>% 
                                       mutate(Total = A+C+G+T+N)
position_base_content_w <- position_base_content_w %>% 
                                        mutate(pcT = T*100/Total) %>% 
                                        mutate(pcA = A*100/Total) %>% 
                                        mutate(pcC = C*100/Total) %>% 
                                        mutate(pcG = G*100/Total) %>% 
                                        mutate(pcN = N*100/Total) %>% 
                                        mutate(sumPC = pcT+pcA+pcC+pcG+pcN)

```




```{r, echo =FALSE, message=FALSE}
#heatmap
p <- ggplot(position_base_content_w, aes(ReadPos, Sample, fill= pcT)) + 
  geom_tile() +
  scale_fill_gradient(low="white", high="red4")+ 
  labs(title ="Dragen-QC: Per-Base position sequence Content; %T")+
  theme(axis.text.y = element_blank()) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "grey90"))

  

ggplotly(p)

```
```{r, echo =FALSE, message=FALSE}
#heatmap
p <- ggplot(position_base_content_w, aes(ReadPos, Sample, fill= pcA)) + 
  geom_tile() +
  scale_fill_gradient(low="white", high="yellow4")+ 
  labs(title ="Dragen-QC: Per-Base position sequence Content; %A")+
  theme(axis.text.y = element_blank()) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "grey90"))

ggplotly(p)

```

```{r, echo =FALSE, message=FALSE}
#heatmap
p <- ggplot(position_base_content_w, aes(ReadPos, Sample, fill= pcC)) + 
  geom_tile() +
  scale_fill_gradient(low="white", high="green4")+ 
  labs(title ="Dragen-QC: Per-Base position sequence Content; %C")+
  theme(axis.text.y = element_blank()) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "grey90"))

ggplotly(p)

```

```{r, echo =FALSE, message=FALSE}
#heatmap
p <- ggplot(position_base_content_w, aes(ReadPos, Sample, fill= pcG)) + 
  geom_tile() +
  scale_fill_gradient(low="white", high="blue4")+ 
  labs(title ="Dragen-QC: Per-Base position sequence Content; %G")+
  theme(axis.text.y = element_blank()) +
  theme(panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "grey90"))

ggplotly(p)

```

```{r, echo =FALSE, message=FALSE}
#heatmap
p <- ggplot(position_base_content_w, aes(ReadPos, Sample, fill= pcN)) + 
  geom_tile() +
  scale_fill_gradient(low="white", high="black")+ 
  labs(title ="Dragen-QC: Per-Base position sequence Content; %N")+
  theme(axis.text.y = element_blank()) +
  theme(panel.background = element_rect(fill = "white"),
       panel.grid.major = element_line(colour = "grey90"))

ggplotly(p)

```

## 4- Read length

Total number of reads with each observed length. 

```{r, echo =FALSE, message=FALSE, results='hide'}
read_lengths <- All_files %>% filter(V1 == "READ LENGTHS")
head(read_lengths)
read_lengths$V3 <- read_lengths$V3 %>% str_remove("bp Length Reads")

read_lengths$V3 <- as.numeric(read_lengths$V3)
names(read_lengths)[names(read_lengths) == 'V3'] <- 'Length'
names(read_lengths)[names(read_lengths) == 'V4'] <- 'ReadCount'
head(read_lengths)
```

```{r, echo =FALSE, message=FALSE}
p <-ggplot(data =read_lengths,  aes(x= Length, y=ReadCount, color = Sample )) +
    labs(title ="Dragen-QC: Sequence Length Distribution", y= "Read Count", x = "Sequence Length (bp)") +
    geom_line() +
    theme(legend.position = "none")  +
    #theme(axis.text.x = element_text(angle=45)) +
    scale_y_continuous(labels = scales::label_comma() ) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
    theme_minimal()

ggplotly(p) %>% layout(showlegend = FALSE)
```





## 5- Read GC Content; Per-Sequence GC Content

Total number of reads with each GC content percentile between 0% and 100%.


```{r, echo =FALSE, message=FALSE, results='hide'}
read_gc_content <- All_files %>% filter(V1 == "READ GC CONTENT")
head(read_gc_content, n=4)

read_gc_content$V3 <- read_gc_content$V3 %>% str_remove("% GC Reads") %>% 
                                             as.numeric()
head(read_gc_content, n=4)

read_gc_content <- read_gc_content %>% 
                   group_by(Sample) %>% 
                   mutate(Total_read = sum(V4)) 

read_gc_content <- read_gc_content %>% 
                   group_by(Sample) %>% 
                   mutate(PercentRead = (V4*100)/Total_read)  %>% 
                   mutate(tt = sum(PercentRead))

names(read_gc_content)[names(read_gc_content) == 'V3'] <- 'PercentGC'
names(read_gc_content)[names(read_gc_content) == 'V4'] <- 'ReadCount'
head(read_gc_content)
```

```{r,echo =FALSE, message=FALSE}
p <-ggplot(data =read_gc_content,  aes(x= PercentGC, y=ReadCount, color = Sample )) +
    labs(title ="Dragen-QC: Per-Sequence GC Content", y= "Count", x = "% GC") +
    geom_line() +
    theme(legend.position = "none")  +
    #theme(axis.text.x = element_text(angle=45)) +
    scale_y_continuous(labels = scales::label_comma() ) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +
    theme_minimal()

ggplotly(p) %>% layout(showlegend = FALSE)
```

```{r, echo =FALSE, message=FALSE}
p <-ggplot(data =read_gc_content,  aes(x= PercentGC, y=PercentRead, color = Sample )) +
    labs(title ="Dragen-QC: Per-Sequence GC Content", y= "% Reads", x = "% GC") +
    geom_line() +
    theme(legend.position = "none")  +
    #theme(axis.text.x = element_text(angle=45)) +
    #scale_y_continuous(labels = scales::label_percent(scale = 100) ) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 15)) +   
    theme_minimal()

ggplotly(p) %>% layout(showlegend = FALSE)

```



## 6- Read GC Content Quality; Average mean quality for reads by GC%

Average Phred-scale read mean quality for reads with each GC content percentile between 0% and 100%.

```{r, echo =FALSE, message=FALSE, results='hide'}
read_gc_content_quality <- All_files %>% filter(V1 == "READ GC CONTENT QUALITY")

read_gc_content_quality$V3 <- read_gc_content_quality$V3 %>% str_remove("% GC Reads Average Quality") %>% 
                                                            as.numeric()

names(read_gc_content_quality)[names(read_gc_content_quality) == 'V3'] <- 'PercentGC'
names(read_gc_content_quality)[names(read_gc_content_quality) == 'V4'] <- 'PhredScore'

head(read_gc_content_quality)

```

```{r, echo =FALSE, message=FALSE}

p <- ggplot(data =read_gc_content_quality,  aes(x= PercentGC, y=PhredScore, color = Sample )) +
    labs(title ="Dragen-QC: Average mean quality for reads vs GC% ", y= "Phred Quality Score", x = "% GC content") +
    geom_line() +
    theme(legend.position = "none")  +
    scale_y_continuous(breaks = scales::pretty_breaks(n = 20)) +
    theme_minimal()

ggplotly(p) %>% layout(showlegend = FALSE)

```



## 7- Sequence Positions; Cumulative Adapter Content

Number of times an adapter or other kmer sequence is found, starting at a given position in the input reads. Sequences are listed first in the metric description in quotes. Locations are listed second and can be either specific positions or ranges.

From the Illumina Dragen manual:  
"DRAGEN contains a novel lossless soft-trimming mode. In soft-trimming mode, reads are mapped as though they had been trimmed, but no bases are removed. The intention of soft trimming is to suppress systematic mismapping of reads containing trimmable artifacts, such as Poly-G artifacts, from getting mapped to reference G homopolymers or adapter sequences getting mapped to matching reference loci, without actually losing the trimmed bases in aligned output."


```{r, echo =FALSE, message=FALSE, results='hide'}
sequence_positions <- All_files %>% filter(V1 == "SEQUENCE POSITIONS")

sequence_positions_total <- sequence_positions %>% filter(V3 == "'AGATCGGAAGAG' Total Sequence Starts")  %>% 
                                             select("Sample", "V4")
head(sequence_positions)
```

```{r, echo =FALSE, message=FALSE, results='hide'}
colnames(sequence_positions_total)[names(sequence_positions_total) == "V4"] <- "Total"
sequence_positions <- sequence_positions %>%  filter(V4 != "'AGATCGGAAGAG' Total Sequence Starts")
  
sequence_positions <- sequence_positions %>% left_join(sequence_positions_total)
```


```{r, echo =FALSE, message=FALSE, results='hide', warning = FALSE}
sequence_positions$V3 <- sequence_positions$V3 %>% str_remove("'AGATCGGAAGAG' ") %>% 
                                                   str_remove("bp Starts") %>% 
                                                   as.numeric()
sequence_positions <- sequence_positions %>% group_by(Sample) %>%  
                                             mutate(csum = cumsum(V4)) %>% 
                                             mutate(Total_read = sum(V4)) %>% 
                                             mutate(perct = csum*100/Total_read)
names(sequence_positions)[names(sequence_positions) == 'V3'] <- 'SequencePosition'
head(sequence_positions) 
```




```{r, echo =FALSE, message=FALSE}

p <- ggplot(data =sequence_positions,  aes(x= SequencePosition, y=perct, color = Sample )) +
    labs(title ="Dragen-QC: Cumulative Adapter Content", y= "% of sequences", x = "Sequence Position (bp)") +
    geom_line() +
    theme(legend.position = "none") + 
    theme_minimal()

ggplotly(p) %>% layout(showlegend = FALSE)

```




## 8- Positional Quality

Phred-scale quality value for bases at a given location and a given quantile of the distribution. Locations are listed first and can be either specific positions or ranges. Quantiles are listed second and can be any whole integer 0–100.  
  
This plot represent the same type of plot as the Box and Whisker plot generated by [fastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc). The major difference is that here we plot all values from fastqc output and not just min, max and the interquartile range.
For samples with low phred score, the plot will become darker. The brighter the plot, the better the overall score.   

```{r, echo =FALSE, message=FALSE, results='hide'}
position_qual <- All_files %>% filter(V1 == "POSITIONAL QUALITY")
head(position_qual, n= 20)
```


```{r, echo =FALSE, message=FALSE, results='hide'}
position_qual$V3 <- position_qual$V3 %>% str_remove("% Quantile QV")

position_qual <- position_qual %>% separate(V3, c("Nm", "ReadPos", "PctQuantile"))
names(position_qual)[names(position_qual) == 'V4'] <- 'PhredScore'
head(position_qual, n= 20)
position_qual$ReadPos <- as.numeric(position_qual$ReadPos)
position_qual$PctQuantile <- factor(position_qual$PctQuantile, levels = rev(c("99", "98", "95", "90", "75", "50",  "25", "10", "5", "2",  "1")))
position_qual$PctQuantile <- factor(position_qual$PctQuantile, levels=rev(levels(position_qual$PctQuantile)))
head(position_qual, n= 20)
levels(position_qual$PctQuantile)

```


```{r, echo =FALSE, message=FALSE, results='hide'}
# Single plot as example, loop plot for all samples below
colors <- rev(c("#FFF5EB", "#FEE9D3", "#FDD8B3", "#FDC28B", "#FDA761", "#FD8D3C",
            "#F3701B", "#E25508", "#C44001", "#9E3203", "#7F2704"))

p <- ggplot(position_qual, aes(x=ReadPos, y=PhredScore, fill=PctQuantile)) + 
    geom_area(position = "identity")+
    scale_fill_manual(values=colors)+
    scale_x_continuous(breaks = scales::pretty_breaks(n = 20)) +
    scale_y_continuous(breaks=seq(0,37,1)) +
    theme(panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(colour = "grey90"))


ggplotly(p)
```




```{r, echo =FALSE, message=FALSE, results = "asis"}
colors <- rev(c("#FFF5EB", "#FEE9D3", "#FDD8B3", "#FDC28B", "#FDA761", "#FD8D3C",
            "#F3701B", "#E25508", "#C44001", "#9E3203", "#7F2704"))

Sample_list <- list()

for  ( i in unique(position_qual$Sample)) {
  Sample_nr <- paste("Sample",i)
  Sample_filter <- filter(position_qual, Sample == i)
#  print(Sample_filter)
  
  plot <- ggplot(Sample_filter, aes(x=ReadPos, y=PhredScore, fill=PctQuantile)) + 
        geom_area(position = "identity")+
        ggtitle(Sample_nr)+
        scale_fill_manual(values=colors)+
        scale_x_continuous(breaks = scales::pretty_breaks(n = 20)) +
        scale_y_continuous(breaks=seq(0,37,1)) +
        theme(panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "grey90"))
  
  
  Sample_list[[i]] = plot
  print(plot)
  
  
}



```




```{r}
sessionInfo()
```













